---
title: "Grouping animal data by domain"
subtitle: "How to store and display tidy data measured at different scales"

execute: 
  warning: false
  message: false
  cache: false
---

## Introduction

I will demonstrate how disparate data can be stored together and accessed in tidy tables.
For this example, I'm using `R` data tables (`tibbles`) because they are most familiar to me,
but all the same features are available in Python and can be utilized for RuFaS. 

For this example, I have generated a fake data set with nonsense values, but whose structure is 
illustrative. This dataset pertains to cows, their pens, and their feed rations. The most important
aspect of these data is that they are measured across different observation units (cows, pens, and feed). 
So, in order to differentiate them, they are grouped by `domain`. In this case the domains are:

- `CowMeasurements`: measurements of individual cows through time.
- `PenInfo`: data about the pens that the individual cows are housed in
- `RationInfo`: data about ration formulations that are fed to the cow
- `CowVariableLookup`: metadata about the different measurements taken from individual cows.

```{r}
library(tidyverse)
grouped_data = readRDS("data/example_cow_dataset.rds")
```

@tbl-nested shows a compact representation of the storage object that contains these data, which is very 
similar to how RuFaS' `OutputManager` (`OM`) currently stores data. The `OM` data is grouped by
`info_map` variables, but here we only include `domain` for simplicity. Also, the `OM` currently 
stores the groups of data as dictionaries, but here they are stored as tables (called "tibbles").

```{r}
#| label: tbl-nested
#| tbl-cap: |
#|   Nested data, grouped by domain. Each element of the data column is a sub-table, with their dimensions 
#|   displayed as [rows x columns].
#| cache: true
grouped_data %>% print()
```

### Cow data

With this grouping structure, we are able to easily extract and "tidy" data that belong together 
(i.e., have the same observation units). So, if we wanted to look at the first few rows 
(of about 1 million) of the `CowMeasurements` data set -- in a tidy format -- we can do so easily. 
@tbl-cow-data shows the first 10 observations of a single cow. note that 10 variables were measured
for each cow (in addition to grouping variables like `pen_id` and `ration_id`).

```{r}
#| label: tbl-cow-data
#| tbl-cap: |
#|   Individual cow data (first 10 observations)
#| echo: true

# extract the data
cow_data <- grouped_data %>% 
  filter(domain == "CowMeasurements") %>% # get only this domain
  pull(data) %>% # extract the data column
  bind_rows() %>% # stack into a data frame
  mutate(across(c(var_1:var_10), ~round(.x, 2))) # round measurements to 2 decimals
  
# display the data
cow_data %>% 
  select(-c("domain")) %>% # drop the invariant "domain" column
  arrange(pen_id, year, cow_id) %>% # sort by pen id, then year, then cow
  head(n = 10) # show first 10 rows
```

and if we wanted a more digestible table, we could aggregate the cow data. @tbl-cow-aggregate demonstrates such an
aggregation summary. For each individual cow, it shows how many years they were observed, the total number
of times they were observed, and the average values for all measured variables across observations. 

```{r}
#| label: tbl-cow-aggregate
#| tbl-cap: |
#|   Aggregated cow data (first 10 individuals)
#| echo: true

# Summarize the measurements for each cow within a pen
cow_data %>% 
  group_by(pen_id, cow_id) %>% # aggregate by pen and cow
  summarize(
    n_years = length(unique(year)), # number of years observed
    n_obs = n(), # count how many observations of this individual
    across(c(var_1:var_10), ~round(mean(.x, na.rm = TRUE), 2)), # avg measured variables by group
    .groups = "drop" # remove the grouping after
  ) %>% 
  head(n = 10) # only show first 10 cows
```

### Variable lookup (units)

One of the `domain` groups of our data storage object is called `CowVariableLookup`. This group is metadata about
the measurements we took from our cows. @tbl-var-lookup shows these metadata in full. Its purpose
is to provide additional info about the variables from the `CowMeasurements` data (including `units`) of the
variables. Importantly, the `var_name` column corresponds to the variable names in that data set.

```{r}
#| label: tbl-var-lookup
#| tbl-cap: "Cow measurement variable lookup table and metadatas"
#| echo: true

# extract the data
grouped_data %>% 
  filter(domain == "CowVariableLookup") %>% # get only this domain
  pull(data) %>% # extract the data column
  bind_rows() # stack into a data frame
```

## Pen and Ration Data

Similarly, the data in the `PenInfo` domain (@tbl-pen-info) pertain to data collected at the pen level. Note that
the `pen_id` column contains the unique identifiers of each pen. The same values are referenced by the
`CowMeasurements` data set. So if we ever need information about about the pen a cow was staying in, we can look it up
or join the tables on this common `pen_id` column.

```{r}
#| label: tbl-pen-info
#| tbl-cap: "Pen-specific data"
#| echo: true

# extract the pen data
grouped_data %>% 
  filter(domain == "PenInfo") %>% # get only this domain
  pull(data) %>% # extract the data column
  bind_rows()
```

Our last example domain is `RationInfo` (@tbl-ration-info), which contains information about the different rations 
that cows were fed. As with the pen data, the `ration_id` column is referenced by the `CowMeasurements` dataset and
can be used to join those data with the ration information.

```{r}
#| label: tbl-ration-info
#| tbl-cap: "ration data"
#| echo: true

# extract the pen data
grouped_data %>% 
  filter(domain == "RationInfo") %>% # get only this domain
  pull(data) %>% # extract the data column
  bind_rows()
```

## Conclusions

The solution outlined here, grouping data by `domain`, allows us to easily store, access, and display data
taken from entirely different observational units and/or aspects of the farm. Each table produced during this
example is reasonably compact, without being overly ragged (i.e., because they don't include variables that aren't 
relevant to the respective observations) and while adhering to the tidy paradigm.
